Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2023-01-03 sports_analytics.Rmd
ctv access date; close #50

While @Lopez2018 use betting market data to model team strengths, they do not directly address strategies for betting or inefficiencies in betting markets. 
Early work by @gandar1988testing examines the rationality of NFL betting markets and concludes that statistical tests fail to reject the hypothesis of rationality.
Related work such as @lacey1990estimation and @Boulier2006 explores the efficiency of NFL betting markets in the mid-1980s and late-1990s, respectively.
Neither paper finds strong evidence for inefficiencies in the markets. 
@boulier2003predicting compare the predictive performance of power rankings and media experts to the betting market for NFL games and found that the betting market is the best for predicting winners. 
@lopez2015building show that betting market data was most useful in predicting men's college basketball tournament outcomes.

@@

Analytical work in sports requires facility with an ever-changing set of computational tools for working with data. 
Sources of authoritative data about sports are myriad, and are too numerous to list here. 
Software tools for sports analytics are similarly numerous. 
For R, we maintain a [CRAN Task View for Sports Analytics](https://CRAN.R-project.org/view=SportsAnalytics) that catalogs R packages published on the Comprehensive R Archive Network (CRAN) and organizes them by sport [@baumer2022ctv].
Table \ref{tab:ctv} provides an overview of the currently available sport-specific CRAN packages.
@casals2022systematic offers a systematic review of sport-related packages on CRAN.
Further, a more general collection of software tools is being curated by the SportsDataverse initiative [@gilani2022sportsdataverse].

```{r}
tibble(
  "American Football" = "nflverse, nflfastR, nflreadr, nfl4th, nflseedR,\n nflplotR, NFLSimulatoR, fflr, ffscrapr, ffsimulator, gsisdecoder, cfbfastR",
  "Association Football (Soccer)" = "worldfootballR, engsoccerdata, socceR, ggsoccer, footballpenaltiesBL, footBayes, itscalledsoccer, FPLdata, EUfootball",
  "Australian Rules Football" = "fitzRoy",
  "Baseball/Softball" = "Lahman, retrosheet, pitchRx, mlbstats, baseballDBR, baseballr, runexp",
  "Basketball" = "BAwiR, AdvancedBasketballStats, uncmbb, BasketballAnalyzeR, NBAloveR, wehoop, hoopR, toRvik",
  "Chess" = "chess, stockfish, bigchess, rchess, chessR",
  "Cricket" = "yorkr, cricketr, cricketdata, howzatR",
  "Esports" = "CSGo, rbedrock, ROpenData, opendotaR, RDota2",
  "GPS Activity Tracking" = "trackeR, trackeRapp, rStrava",
  "Hockey" = "hockeyR, NHLData, nhlapi, nhlscrape, fastRhockey",
  "Swimming" = "SwimmeR",
  "Track and Field" = "combinedevents, JumpeR",
  "Volleyball" = "volleystat"
) |> 
  pivot_longer(
    everything(),
    names_to = "Sport",
    values_to = "Packages"
  ) |>
  transmute(
    Sport,
    "Number of Packages" = str_count(Packages, ",") + 1,
    "List of Packages" = Packages
  ) |>
  arrange(desc(`Number of Packages`)) |>
  kable(
    booktabs = TRUE,
    linesep = "",
    caption = "A summary of sport-specific packages available on the Comprehensive R Archive Network (CRAN). While the major North American sports dominate the list, perhaps the fastest-growing collection is for esports. \\label{tab:ctv}"
  ) |> 
  column_spec(3, width = "7cm")
```

>>>>

While @Lopez2018 use betting market data to model team strengths, they do not directly address strategies for betting or inefficiencies in betting markets. 
Early work by @gandar1988testing examine the rationality of NFL betting markets and concludes that statistical tests fail to reject the hypothesis of rationality.
Related work such as @lacey1990estimation and @Boulier2006 explores the efficiency of NFL betting markets in the mid-1980s and late-1990s, respectively.
Neither paper finds strong evidence for inefficiencies in the markets. 
@boulier2003predicting compare the predictive performance of power rankings and media experts to the betting market for NFL games and found that the betting market is the best for predicting winners. 
@lopez2015building show that betting market data was most useful in predicting men's college basketball tournament outcomes.

@@

Analytical work in sports requires facility with an ever-changing set of computational tools for working with data. 
Sources of authoritative data about sports are myriad, and are too numerous to list here. 
Software tools for sports analytics are similarly numerous. 
For R, we maintain a [CRAN Task View for Sports Analytics](https://CRAN.R-project.org/view=SportsAnalytics) that catalogs R packages published on the Comprehensive R Archive Network (CRAN) and organizes them by sport [@baumer2022ctv].
Table \ref{tab:ctv} provides an overview of the currently available sport-specific CRAN packages.
Recently, @casals2022systematic offer a systematic review of sport-related packages on CRAN.
Further, a more general collection of software tools is being curated by the SportsDataverse initiative [@gilani2022sportsdataverse].

```{r}
tibble(
  "American Football" = "nflverse, nflfastR, nflreadr, nfl4th, nflseedR,\n nflplotR, NFLSimulatoR, fflr, ffscrapr, ffsimulator, gsisdecoder, cfbfastR",
  "Association Football (Soccer)" = "worldfootballR, engsoccerdata, socceR, ggsoccer, footballpenaltiesBL, footBayes, itscalledsoccer, FPLdata, EUfootball",
  "Australian Rules Football" = "fitzRoy",
  "Baseball/Softball" = "Lahman, retrosheet, pitchRx, mlbstats, baseballDBR, baseballr, runexp",
  "Basketball" = "BAwiR, AdvancedBasketballStats, uncmbb, BasketballAnalyzeR, NBAloveR, wehoop, hoopR, toRvik",
  "Chess" = "chess, stockfish, bigchess, rchess, chessR",
  "Cricket" = "yorkr, cricketr, cricketdata, howzatR",
  "Esports" = "CSGo, rbedrock, ROpenData, opendotaR, RDota2",
  "GPS Activity Tracking" = "trackeR, trackeRapp, rStrava",
  "Hockey" = "hockeyR, NHLData, nhlapi, nhlscrape, fastRhockey",
  "Swimming" = "SwimmeR",
  "Track and Field" = "combinedevents, JumpeR",
  "Volleyball" = "volleystat"
) |> 
  pivot_longer(
    everything(),
    names_to = "Sport",
    values_to = "Packages"
  ) |>
  transmute(
    Sport,
    "Number of Packages" = str_count(Packages, ",") + 1,
    "List of Packages" = Packages
  ) |>
  arrange(desc(`Number of Packages`)) |>
  kable(
    booktabs = TRUE,
    linesep = "",
    caption = "A summary of sport-specific packages available on the Comprehensive R Archive Network (CRAN) as of October 16, 2022. While the major North American sports dominate the list, perhaps the fastest-growing collection is for esports. \\label{tab:ctv}"
  ) |> 
  column_spec(3, width = "7cm")
```
